---
title: "Differential abundance analysis in R"
execute:
  warning: false
  message: false
output:
  html_document: default
editor: 
  markdown: 
    wrap: 72
---

## Introduction

In this part, we continue to perform differential abundance analysis on
the processed imzML from Section 1 to find precursors that present high
intensity in the microbial culture region. First we apply a non-spatial
linear fixed effects model, then explore modeling with spatial
correlation. In addition, we also contrast differential analysis on
features with and without ion mobility.

## Package setup

We will use `Cardinal` for data import and normalization, `spaMM` for
differential analysis and `ggplot2` for visualization. If the packages
are not yet installed, run following code in a R session:

```{r installation}
#| eval: false
#| echo: true
#| warning: false
#| message: false
#| collapse: true
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Cardinal")
package.install("spaMM")
package.install("ggplot2")
```

Then load the packages:

```{r setup}
library(Cardinal)
library(spaMM)
library(ggplot2)
```

## Load processed dataset

First we load the processd imzML from TIMSImaging into Cardinal:

```{r load data}
msa <- readMSIData("D:\\dataset\\Laura_Gordon\\laura_gordon.imzML", as="MSImagingArrays", extraArrays=c(mobility="MS:1003006"))
mse <- convertMSImagingArrays2Experiment(msa)
mse
```

check the regions:

```{r}
image(mse)
```

## Data preparation

Then we label 3 microbioal regions as a group(`culture`) and the matrix
region as the other group(`media`), which is consistent with the
literature.

```{r define region}
coords <- coord(mse)
pixel_label <- ifelse(coords$x > 200 & coords$y > 100, "media", "culture")
pData(mse)$label <- factor(pixel_label)
pData(mse)
```

To reduce the complexity of spatial modeling, we just compare the
*G.arilaitensis*(bottom middle) region(`culture`) and the media/matrix
region(`media`). For targets in following MS2 acquisition, we want to
exclude matrix/media ions and select precursors spatially associated
with the microbioal culture region. Specifically, a desired precursor
should present high intensity in the microbioal culture region and
minimum intensity in the matrix/media region.\

```{r select regions}
mse_subset <- subsetPixels(mse, x>120, y>100)
image(mse_subset)
```

First we try differential abundance analysis on a feature(mz=428.26)
reported in the paper. Extract the ion image as a data frame, with
pixels labeled as either in 'media' or 'culture' region.

```{r create dataframe}
m <- 428.26
i <- findInterval(m, mz(mse_subset))+1
intensity <- spectraData(mse_subset)$intensity[i,]
df <- data.frame(
  intensity = intensity,
  label = pData(mse_subset)$label,
  x = pData(mse_subset)$x,
  y = pData(mse_subset)$y
)
# set the media group as the baseline
df$label <- relevel(df$label, ref="media")
```

Plot its ion image:

```{r}
image(mse_subset, i=i)
```

## Fitting a non-spatial model

Now we fit a non-spatial linear fixed effect model on this feature. Let
$s$ be a pixel position, $k$ be a group label, then the intensity is a
linear response to group label: $$y_s=\mu_k+\varepsilon_s$$ where $y_s$
is the intensity at $s$, the fixed effect $\mu_k$ is the mean intensity
in group $k$, $\varepsilon_s~\mathcal{N}(0, \sigma^2)$ is the random
error at $s$.

```{r non-spatial modeling}
model <- fitme(
    intensity ~ label,
    data = df,
    method = "REML",
)
summary.HLfit(model, details=c(p_value="Wald"))
```

The result shows that ion m/z=428.26 present higher intensity in the
culture region(`labelculture`) than the media region(`intercept`) with
high significance score. However, in the non-spatial modeling the pixels
were considered as independent samples, which is not true(pixels are
highly correlated), the degree of freedom is overstimated and resulted
in extremely small p-vales(type-I error).

## Fitting a spatial model

Next, we try a spatial linear mixed effect model with Matérn
correlation. The model could be expressed as
$$y_s=\mu_k+U_s+\varepsilon_s$$where $y_s$ is the intensity at $s$, the
fixed effect $\mu_k$ is the mean intensity in group $k$, the random
effect $U_s$ is the Matérn random field at $s$, and
$\varepsilon_s \sim \mathcal{N}(0, \sigma^2)$ is the random error at
$s$. The Matérn random field is sampled from
$\mathcal{N}(\mathbf{0}, \sigma^2\mathbf{V})$, that the correlation
matrix $\mathbf{V}$ is $$
V_{ij}=
\mathrm{corr}(d_{ij})
=
\frac{1}{2^{\nu-1}\Gamma(\nu)}
\left( \frac{d}{\rho} \right)^{\nu}
K_{\nu}\!\left( \frac{d}{\rho} \right)
$$where $d$ is the Euclidean distance between pixel $i$ and $j$,
$\Gamma$ is the Gamma function, $K_{\nu}$ is the modified Bessel
function of the second kind with order $\nu$, and $\rho$ and $\nu$ are
parameters. Here we tried the commonly used special case with $\nu=0.5$,
that the correlation decays exponentially: $$
\mathrm{corr}(d_{ij})=\exp\!\left(-\frac{d}{\rho}\right)
$$

```{r spatial modeling}
model <- fitme(
    intensity ~ label + Matern(1|x+y),
    data = df,
    fixed = list(nu=0.5),
    method = "REML",
)
summary.HLfit(model, details=c(p_value="Wald"))
```

The spatial model also shows the differential abundance between two
regions, but with more reasonable p-value.

##Fitting non-spatial and spatial model on all features
Then we can loop over all features by fitting a model on each of them:

```{r batched processing}
nonspatial_de <-function(mse){
  mz_values <- mz(mse)
  fit_results <- lapply(seq_along(mz_values), function(i) {
    
  intensity_i <- spectraData(mse)$intensity[i, ]
  
  # create a subset data frame
  sub_df <- data.frame(
    intensity = intensity_i,
    label = pData(mse)$label,
    x = coord(mse)$x,
    y = coord(mse)$y
  )
  # set the media group as the baseline
  sub_df$label <- relevel(sub_df$label, ref="media")
  # fit a linear fixed effect model
  model <- tryCatch({
    fitme(intensity ~ label,
          data = sub_df,
          method = "REML")
  }, error = function(e) NULL)
  
  # summarize the results
  if (!is.null(model)) {
    #coefs <- fixed.effects(model)
    stats <- summary.HLfit(model, details=c(p_value="Wald"), verbose=FALSE)[['beta_table']]
    
    return(data.frame(
      mz = mz_values[i],
      index = i,
      intercept = stats[1,1],
      label_effect = stats[2,1],
      t_value = stats[2,3],
      p_value = stats[2,4]
    ))
  } else {
    return(NULL)
  }
})
# combine all results
fit_results_df <- do.call(rbind, fit_results)
fit_results_df$intensity <- fit_results_df$intercept+fit_results_df$label_effect
fit_results_df$log2_foldchange <- log2((fit_results_df$intercept+fit_results_df$label_effect)/fit_results_df$intercept)
fit_results_df$neg_log10_p <- pmin(-log10(fit_results_df$p_value), 20)
return(fit_results_df)
}
```

Summarize the results:

```{r}
fit_results_df <- nonspatial_de(mse_subset)
head(fit_results_df, 10)
```

Here is the code for spatial model:

```{r}
#| eval: false
#| echo: true
#| warning: false
#| message: false
#| collapse: true
#| 
spatial_de <- function(mse){
  mz_values <- mz(mse)
  fit_results <- lapply(seq_along(mz_values), function(i) {
    message("Fitting feature", i)
    intensity_i <- spectraData(mse)$intensity[i, ]
  
    sub_df <- data.frame(
      intensity = intensity_i,
      label = pData(mse)$label,
      x = coord(mse)$x,
      y = coord(mse)$y
    )
    sub_df$label <- relevel(sub_df$label, ref="media")
  
    model <- tryCatch({
      fitme(intensity ~ label + Matern(1|x+y),
            data = sub_df,
            fixed = list(nu=0.5),
            control.HLfit = list(algebra="spcorr", NbThreads=8),
            method = "REML")
    }, error = function(e) NULL)
    
    if (!is.null(model)) {
      coefs <- fixed.effects(model)
      stats <- summary.HLfit(model, details=c(p_value="Wald"), verbose=FALSE)[['beta_table']]
      
      return(data.frame(
        mz = mz_values[i],
        intercept = coefs[1],
        label_effect = coefs[2],
        t_value = stats[2,3],
        p_value = stats[2,4]
      ))
    } else {
      return(NULL)
    }
  })
  fit_results_df <- do.call(rbind, fit_results)
  fit_results_df$log2_foldchange <- log2((fit_results_df$intercept+fit_results_df$label_effect)/fit_results_df$intercept)
  fit_results_df$neg_log10_p <- pmin(-log10(fit_results_df$p_value), 20)
  return(fit_results_df)
}
```

It takes hours to run the spatial model on all the features. For
presentation here we load the pre-computed results:

```{r}
fit_results_df_sp <- read.csv("D:\\dataset\\Laura_Gordon\\spatial_model_results.csv")
head(fit_results_df_sp, 10)
```

Now we can visualize the results with volcano plots.

```{r volcano plot}
volcano_plot <- function(df){
  p <- ggplot(df, aes(x = log2_foldchange, y = neg_log10_p)) +
    geom_point(alpha = 0.7) +
    geom_hline(yintercept = -log10(0.05), color = "red", linetype = "dashed") +  # p=0.05 threshold
    geom_vline(xintercept = 0, color = "blue", linetype = "dotted") +
    labs(
      title = "Volcano Plot of m/z Features",
      x = "log2(fold change)",
      y = "-log10(p-value)"
    ) +
    coord_cartesian(xlim = c(-5, 5), ylim=c(0,20)) +
    theme_minimal()
  return(p)
}
```

```{r}
volcano_plot(fit_results_df)
volcano_plot(fit_results_df_sp)
```

The spatial modeling resulted in a more reasonable volcano plot. Points
on the top right(significantly up-expressed in the microbial culture
region with high fold change) are the candidate precursors. For example,

```{r precursor selection}
selected_precursors <- subset(fit_results_df_sp, (log2_foldchange>2) & (p_value<0.05))
selected_precursors <- selected_precursors[order(selected_precursors$neg_log10_p, decreasing = TRUE),]
head(selected_precursors, 10)
```

Plot ion images for some example precursors with differential abundance:

```{r}
indices = as.numeric(row.names(selected_precursors)[1:6])
image(mse_subset, i=indices)
```

## Processing TIMSCONVERT data

In order to show ion mobility also benefits differential abundance
analysis, we contrast the results from processing without ion mobility.

```{r processing without ion mobility}
msa_timsconvert <- readMSIData("D:\\dataset\\Laura_Gordon\\250321_JB182_Pen12.imzML")
mse_timsconvert_binned <- bin(msa_timsconvert, resolution=20, units="ppm")
centroided(mse_timsconvert_binned)<-FALSE
set.seed(42, kind="L'Ecuyer-CMRG")
peaks_timsconvert <- peakProcess(mse_timsconvert_binned, SNR=6, tolerance=400, units="ppm")
```

Then we filtered the features by intensity and got 378 features.

```{r}
peaks_timsconvert <- summarizeFeatures(peaks_timsconvert)
max_intensity <- max(fData(peaks_timsconvert)$mean)
intensity_filter <- fData(peaks_timsconvert)$mean > 0.01*max_intensity
peaks_timsconvert <- subsetFeatures(peaks_timsconvert, select=intensity_filter)
peaks_timsconvert
```

Here is the code to run spatial-aware differential abundance analysis on
these features, for time being we also use pre-computed results.

```{r}
#| eval: false
#| echo: true
#| warning: false
#| message: false
#| collapse: true
#| 
pData(peaks_timsconvert)$label <- factor(pixel_label)
peaks_timsconvert_subset <- subsetPixels(peaks_timsconvert, x>120, y>100)
fit_results_df_timsconvert <- nonspatial_de(peaks_timsconvert_subset)
```

```{r}
fit_results_df_timsconvert <- read.csv("D:\\dataset\\Laura_Gordon\\spatial_timsconvert.csv")
```

Next we find the matching features between with and without ion
mobility. Since there could be multiple isobaric features differentiated
by ion mobility and detected by TIMSImaging, a `mz_noim` value might
match with multiple `mz_im` values.

```{r compare p-values}
mz_im <- fit_results_df_sp$mz
mz_noim <- fit_results_df_timsconvert$mz
idx <- sapply(mz_im, function(v) which.min(abs(mz_noim - v)))
spatialde_results <- data.frame(
  mz_im = mz_im,
  mz_noim = mz_noim[idx],
  
  mob = spectraData(msa)$mobility$'spectrum=1',
  tolerance = (mz_im-mz_noim[idx])/mz_im,
  fc_im = fit_results_df_sp$log2_foldchange,
  fc_noim = fit_results_df_timsconvert$log2_foldchange[idx],
  pvalue_im = fit_results_df_sp$neg_log10_p,
  pvalue_noim = fit_results_df_timsconvert$neg_log10_p[idx]
)
# find features with high fold change
spatialde_results <- subset(spatialde_results, (abs(tolerance)<2e-5) & (fc_im>1))
head(spatialde_results, 10)
```

Now we can contrast the p-values for matched features. The x axis is the
p-value processing without ion mobility, and the y axis is correponding
ion moblity-aware p-value. Most points are above the diagonal, showing
that the ion mobility-aware approach of TIMSImaging produced smaller
p-values for differential abundance analysis.

```{r p-value visualization}
ggplot(spatialde_results, aes(x = pvalue_noim, y = pvalue_im)) +
  geom_point(alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed")+
  labs(
    title = "Contrast of p-values",
    x = "without ion mobility",
    y = "with ion mobility"
  ) +
  coord_fixed(ratio=1, xlim = c(0, 10), ylim=c(0,10))
```

<!--Finally, we can plot example ion images for isobaric features, that ion mobility helps find differential features from its less significant isobars. -->

```{r}
#| eval: false
#| echo: false
#| warning: false
#| message: false
#| collapse: true
#| 
i <- 736
img1 <- image(mse_subset, i=i)
mob1 <- spatialde_results[as.character(i), ]$mob
img1$labels <- paste(img1$labels, "\n1/K_0 = ", sprintf("%.3f", mob1))

m <- 877.0435
i <- findInterval(m, mz(mse_subset))
img2 <- image(mse_subset, i=i)
mob2 <- spectraData(msa)$mobility$'spectrum=1'[i]
img2$labels <- paste(img2$labels, "\n1/K_0 = ", sprintf("%.3f", mob2))

m <- 877.0582
i <- findInterval(m, mz(peaks_timsconvert_subset))
img3 <- image(peaks_timsconvert_subset, i=i)
img3$labels <- paste(img3$labels, "\n(no ion mobility)")

plot(img1)
plot(img2)
plot(img3)
```
