---
title: "TIMSCONVERT comparison"
author: "Yinyue Zhu"
date: "2025-07-22"
output: html_document
---

### Introduction

In this vignette, we will compare TIMSImaging workflow with existing method that ignores ion mobility in terms of ion image quality. TIMSCONVERT is an open-source tool to convert general Bruker raw data into open formats. Specifically, it outputs imzML for MALDI-TIMS-MS imaging data, either with ion mobility or not. However, it does not do any pre-processing.

### Process data from TIMSCONVERT

```{r setup, include=FALSE}
library("Cardinal")
```

Load the imzML data from TIMSCONVERT, which is obtained by running the shell script.

```{r data import}
msa <- readMSIData("D:\\dataset\\Melanie_case_study\\Kidney_MS1_ITO6.imzML")
msa
```

TIMSCONVERT keeped ion mobility in the imzML data, however the peak process functions in Cardinal cannot take advantage of it. Here we bin the spectra to project data points with the same m/z but different ion mobilities together.

```{r}
mse_binned <- bin(msa, resolution=20, units="ppm")
mse<-summarizePixels(mse_binned)
```

To reduce the computation, we only process the kidney tissue region in following steps.

```{r}
kidney <- subsetPixels(mse, x<500)
image(kidney, 'tic')
```

Plot one spectrum, though the data is centroided on m/z dimension by the instrument, it looks more similar with a profile spectrum after projection.

```{r unprocessed spectra visualization}
plot(kidney, i=1234)
```

Treat the data as profile spectra and process it(to reduce features)

```{r peak processing}
kidney <- subsetPixels(mse_binned, x<500)
centroided(kidney)<-FALSE
set.seed(1, kind="L'Ecuyer-CMRG")
#kidney <- normalize(kidney, method='rms')
peaks <- peakProcess(kidney, SNR=3, tolerance=200, units="ppm")
```

### Comparing ion images

Then we load the processed data from TIMSImaging, which is already peak-picked.

```{r TIMSImaging data}
peaks_timsimaging <- readMSIData("D:\\dataset\\Melanie_case_study\\mouse_kidney.imzML")
peaks_timsimaging
```

Normalize on peak-picked data for fair comparison

```{r normalization}
peaks_norm = process(normalize(peaks, method='tic'))
peaks_timsimaging <- process(normalize(peaks_timsimaging, method='tic'))
```

Plot ion images from processsing with/without ion mobility in the same color scale

```{r ion image visualization}
m <- 1198.7
image(peaks_timsimaging, i = findInterval(m, mz(peaks_timsimaging))+1, zlim=c(0, 5))
image(peaks_norm, mz = m, zlim=c(0, 5))
```

The ion images from TIMSCONVERT and processed in Cardinal looks noiser, as well as less contrast between the background and kidney edge.

```{r}
m <- 1443.59
image(peaks_timsimaging, i = findInterval(m, mz(peaks_timsimaging))+1, zlim=c(0, 8))
image(peaks_norm, mz = m, zlim=c(0, 8))
```
